---
title: "Paternity analysis in *Antirrhinum majus*"
author: "Tom Ellis"
date: "31st October 2019"
output:
  ioslides_presentation:
    widescreen: yes
    css: gmi.css
---

<script src="https://ajax.googleapis.com/ajax/libs/jquery/1.12.2/jquery.min.js"></script>

<script>
$(document).ready(function() {
$('slide:not(.title-slide, .backdrop, .segue)').append('<footer></footer>');    
$('footer').attr('label', 'thomas.ellis@gmi.oeaw.ac.at @ellisztamas');

})
</script>

<style>
footer:after {
font-size: 12pt;
content: attr(label);
position: absolute;
bottom: 1px;
right: 60px;
line-height: 1.9;
display: block;
color: #007243
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache=T, cache.lazy=T, fig.align='center', warning = F)
format(Sys.time(), '%d %B, %Y')

library('tidyverse')
```

```{r import-data, include=F}
gps <- read_csv("001.data/002.processed/GPS_positions.csv")

# PATERNITY ANALYSES
# Import parameters on assortment from paternity analysis
mcmc <- read_tsv("004.output/5_parameter_noassortment_repA.txt")
mcmc <- mcmc[seq(500,30000, 10), ]# %>% 
# rename("RR" = "R/R", "Rr" = "R/r", "rr" = "r/r") %>% 
# rename("S_" = "S/+", "ss" = "s/s")
# Import realised mating probabilities between genotypes.
real_ros <- read_csv("004.output/real_mating_rosea.csv")
real_sulf <- read_csv("004.output/real_mating_sulfurea.csv")
# Import mating probabilties under random mating
rand_ros <- read_csv("004.output/random_mating_rosea.csv")
rand_sulf <- read_csv("004.output/random_mating_sulfurea.csv")

# Realised mating events based on posterior mean dispersal
real_mating <- read_csv("004.output/realised_mating_events.csv")
```

# Pollen dispersal kernel
## Generalised exponential distribution

$$ exp[-(\frac{d}{a})^{b}] $$

* $d=$ distance
- $a=$ scale (roughly mean distance travelled)
* $b=$ shape

## Generalised exponential distribution

```{r}
generalised_normal_PDF <- function(x, scale, shape){
  (shape/(2 * scale * gamma(1.0/shape) )) * exp(-(x/scale)**shape)
}

xv <- seq(0, 5, 0.01)

yv_norm <- generalised_normal_PDF(xv, 1, 2)
yv_exp <-  generalised_normal_PDF(xv, 1, 1)
yv_lept <- generalised_normal_PDF(xv, 1, 0.5)

yv_exp <- yv_exp / sum(yv_exp)
yv_norm <- yv_norm / sum(yv_norm)
yv_lept <- yv_lept / sum(yv_lept)

plot(xv, yv_exp, type='l', ylim=c(0,0.015),
     xlab = "Euclidean distance",
     ylab="Probability density",
     lwd=2, col=2)
# points(xv, yv_norm, type='l', lwd=2, col=3)
# points(xv, yv_lept, type='l', lwd=2, col=4)

legend('topright', legend = c("Exponential"), pch=16, col=2)

```

- Shape = 1
- Exponential distribution

## Generalised exponential distribution

```{r}
plot(xv, yv_exp, type='l', ylim=c(0,0.015),
     xlab = "Euclidean distance",
     ylab="Probability density",
     lwd=2, col=2)
points(xv, yv_norm, type='l', lwd=2, col=3)
# points(xv, yv_lept, type='l', lwd=2, col=4)

legend('topright', legend = c("Exponential", "Gaussian"), pch=16, col=2:3)

```

- Shape = 2
- Gaussian distribution

## Generalised exponential distribution

```{r}
plot(xv, yv_exp, type='l', ylim=c(0,0.015),
     xlab = "Euclidean distance",
     ylab="Probability density",
     lwd=2, col=2)
points(xv, yv_norm, type='l', lwd=2, col=3)
points(xv, yv_lept, type='l', lwd=2, col=4)

legend('topright', legend = c("Exponential", "Gaussian", "Leptokurtic"), pch=16, col=2:4)

```

- Shape < 1
- Fat-tailed (leptokurtic)

## Mixture model

$$ \lambda exp[-(\frac{d}{a})^{b}]  + (1- \lambda) \frac{1}{N} $$

* $d=$ distance
* $a=$ scale (roughly mean distance travelled)
* $b=$ shape
* $\lambda=$ mixture parameter

## 4 parameters to infer

$$ \lambda exp[-(\frac{d}{a})^{b}]  + (1- \lambda) \frac{1}{N} $$

* $a=$ scale (roughly mean distance travelled)
* $b=$ shape
* $\lambda=$ mixture parameter
* proportion $\theta$ of unsampled fathers

## Model checking

```{r}
hist(mcmc$lambda,
     xlab="Mixture parameter",
     ylab="Posterior density",
     main="",
     xlim=c(0,1))
```

- Most of the signal is from the 'real' dispersal kernel

## Model checking

```{r}
hist(mcmc$missing,
     xlab="Proportion missing fathers",
     ylab="Posterior density",
     main="",
     xlim=c(0,1))
x <- seq(0,1, 0.01)
yp <- dbeta(x, 3, 20) 
points(x, yp*100, type='l', col=2, lwd=2)

legend('topright', legend = "beta prior", col=2, pch=16, bty='n')
```

- Not much signal about missing fathers

# Dispersal

## Dispersal is leptokurtic

```{r}
hist(mcmc$shape,
     xlab="Shape parameter",
     ylab="Posterior density",
     main="",
     xlim=c(0,1))
x <- seq(0,1, 0.01)
yp <- dbeta(x, 1.1, 1.1) 
points(x, yp*100, type='l', col=2, lwd=2)
legend('topright', legend = "beta prior", col=2, pch=16, bty='n')

```

* shape < 1
* Dispersal mostly short range
* Rare long-distance events

## Distances are plausible 

<div style="float: left; width: 50%;">

```{r, fig.width=4}
hist(mcmc$mean_dist,
     xlab="Mean dispersal distance (m)",
     ylab="Posterior density",
     main="",
     freq=F,
     xlim=c(0,250),
     breaks=20
)
x <- seq(0,250, 0.01)
yp <- dgamma(x, shape=2, scale=200) 
points(x, yp*5, type='l', col=2, lwd=2)
legend('topright', legend = "gamma prior", col=2, pch=16, bty='n')

```

</div>

<div style="float: right; width: 50%;">

* Mean dispersal:
$$ \frac{a\Gamma(2/b) } {1/b} $$
* $a=$scale; $b=$ shape

</div>

## Posterior-mean kernel shape

```{r obs-dispersal}
xv <- 1:1000
yv <- generalised_normal_PDF(xv, scale =mean(mcmc$scale), shape = mean(mcmc$shape))
mss <- sum(real_mating$prob)

library(plotrix)
weighted.hist(real_mating$distance,
              w = real_mating$prob,
              breaks = seq(0,1000, 10),
              freq=T,)
points(xv, yv*20000, type='l', col=2, lwd=2)

legend('topright', c("Mean kernel shape", "'observed' distances"), pch=16, col=2:1)
```

- $\bar{a}$ = `r mean(mcmc$scale)`; $\bar{b}$ = `r mean(mcmc$shape)`
- bars: observed distances weighted by probability

## Posterior-mean kernel shape

```{r, ref.label="obs-dispersal"}
```

- 'Hump' at 200-400m
- Fathers from the upper road

## Posterior-mean kernel shape

```{r, ref.label="obs-dispersal"}
```

- Tail probably inflated
- Sample densities uneven

# Mating for flower colour
## Observed vs. expected mating

- For each draw in the MCMC: 
  - **Observed**: fathers drawn based on dispersal plus genetics
  - **Expected**: fathers drawn based on dispersal only

```{r}
ros_mating <- vector('list', 3)

for(g in c('R/R', "R/r", "r/r")){
  o <- real_ros %>%
    filter(maternal == g) %>% 
    select(- i, -maternal)
  
  e <- rand_ros %>% 
    filter(maternal == g) %>% 
    select(- i, -maternal)
  
  ros_mating[[g]] <- data.frame(
    maternal = g,
    paternal = colnames(o),
    mean  = colMeans(o-e),
    lower = apply((o-e),2, quantile, 0.025),
    upper = apply((o-e),2, quantile, 0.975)
  )
}
ros_mating <- do.call('rbind', ros_mating)

```

```{r}
sulf_mating <- vector('list', 2)

for(g in c('S/+', 's/s')){
  o <- real_sulf %>%
    filter(maternal == g) %>% 
    select(- i, -maternal)
  
  e <- rand_sulf %>% 
    filter(maternal == g) %>% 
    select(- i, -maternal)
  
  sulf_mating[[g]] <- data.frame(
    maternal = g,
    paternal = colnames(o),
    mean  = colMeans(o-e),
    lower = apply((o-e),2, quantile, 0.025),
    upper = apply((o-e),2, quantile, 0.975)
  )
}
sulf_mating <- do.call('rbind', sulf_mating)
```

## Pollen swamping by magenta plants 


<div style="float: left; width: 50%;">

```{r, fig.width=11/2.54}
os <- 0.1

xv <- rep(1:3, each=3) + rep(c(-os, 0, os), 3)

plot(c(0,4), c(-0.5,0.5), type='n',
     xlab="Maternal genotype",
     ylab="Relative prob. receiving pollen",
     xlim=c(0.5, 3.5),
     axes=F)
axis(1, 1:3, labels = c("R/R", "R/r", "r/r"))
legend('bottomright', legend=c("R/R donors", "R/r donors", "r/r donors"), pt.bg=c('magenta', 'pink', 'white'), pch=21)
axis(2)
box()

abline(0,0, lty=2)
segments(xv, ros_mating$lower, xv, ros_mating$upper, lwd=2)
points(xv, ros_mating$mean, pch=21, bg=c('magenta', 'pink', 'white'), lwd=1)

```

- Observed-expected
- +/- credible intervals

</div>

<div style="float: right; width: 50%;">

```{r fig.width=11/2.54, fig.height=5, echo=FALSE}
library(png)
library(grid)
img <- readPNG("figures/ros_sulf_examples.png")
grid.raster(img)
```


</div>

## Pollen swamping by non-yellow plants 


<div style="float: left; width: 50%;">

```{r, fig.width=11/2.54}
os <- 0.1

xv <- rep(1:2, each=2) + rep(c(-os, os), 2)

plot(c(0,3), c(-0.5,0.5), type='n',
     xlab="Maternal genotype",
     ylab="Relative prob. receiving pollen",
     xlim=c(0.5, 2.5),
     axes=F)
axis(1, 1:2, labels = c("S/+", "s/s"))
legend('bottomright', legend=c("S/+ donors", "s/s donors"), pt.bg=c('white', 'gold'), pch=21)
axis(2)
box()

abline(0,0, lty=2)
segments(xv, sulf_mating$lower, xv, sulf_mating$upper, lwd=2)
points(xv, sulf_mating$mean, pch=21, bg=c('white', 'gold'), lwd=1)

```

- Observed-expected
- +/- credible intervals

</div>

<div style="float: right; width: 50%;">

```{r fig.width=11/2.54, fig.height=5, echo=FALSE}
library(png)
library(grid)
img <- readPNG("figures/ros_sulf_examples.png")
grid.raster(img)
```


</div>

## Male fitness

- *A. m. pseudomajus*-like plants dominate pollen export, but:
  - Other experiments suggest pollinators show no preference
  - No differences in pollen fertility

## Male fitness

- *A. m. pseudomajus*-like plants dominate pollen export, but:
  - Other experiments suggest pollinators show no preference
  - No differences in pollen fertility
- Samples happen to be close to where *A. m. pseudomajus* shows high fitness.

## Male fitness

- *A. m. pseudomajus*-like plants dominate pollen export, but:
  - Other experiments suggest pollinators show no preference
  - No differences in pollen fertility
- Samples happen to be close to where *A. m. pseudomajus* shows high fitness.
- Consistent with cline discordance

## Male fitness

- *A. m. pseudomajus*-like plants dominate pollen export, but:
  - Other experiments suggest pollinators show no preference
  - No differences in pollen fertility
- Samples happen to be close to where *A. m. pseudomajus* shows high fitness.
- Consistent with cline discordance
- How much selection can this explain?

## Conclusions

- Joint analysis of sibship structure, paternity and dispersal
- Order of magnitude larger than previous studies

## Conclusions

- Joint analysis of sibship structure, paternity and dispersal
- Order of magnitude larger than previous studies
- Pollen moves 50-80m on average
- Long tail of long-distance mating events

## Conclusions

- Joint analysis of sibship structure, paternity and dispersal
- Order of magnitude larger than previous studies
- Pollen moves 50-80m on average
- Long tail of long-distance mating events
- Magenta *A. m. pseudomajus* plants show high male fitness

## Acknowledgements

-Switch back